---
title: "EuroBioC2025 OSTA Workshop"
author: Estella Yixing Dong^[estelladong729@gmail.com]
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EuroBioC2025 OSTA Workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Workflow: Visium HD cell-level

Authors: Yixing E. Dong[^1], ABC[^2]. <br/> Last modified: 23 July, 2025.

[^1]: University of Lausanne, Lausanne, Switzerland

[^2]: ABC

## Instructor name and contact information
-   Yixing Estella Dong (estella.yixing.dong@gmail.com)

## Workshop Description

In this instructor-led live demo, we analyse Visium HD data segmented to cells, demonstrating use of `SpatialExperiment` and `sf` classes in R to import, organise, quality control, clustering, marker gene identificaiton, and label transfer to annotate the data. Spatial transcriptomics data with cell types annotated can reveal key insights with neighbourhood enrichment and spatial statistics, which will be demonstrated in a second half of the workshop. The complete analysis offered by existing tools highlights the ease with which researchers can turn the raw counts from a Visium HD experiment into biological insights using Bioconductor and CRAN. The complete workflow is available at <TODO> .

### Pre-requisites

-   Basic knowledge of spatial transcriptomics
-   Basic knowledge of R syntax
-   Familiarity with `SpatialExperiment` classes and object manipulation
-   Familiarity with spatial `sf` classes and object manipulation

Additional reading:

-   Please have a read of [Visium HD file structure](https://lmweber.org/OSTA/pages/bkg-importing.html) and [bin-wise Visium HD analysis](https://lmweber.org/OSTA/pages/seq-workflow-visium-hd.html) in the OSTA book

### Participation

Participants can watch the live demo, or may prefer to follow the demonstration by bringing their laptops along. To follow the analysis on their own laptops, participants need to install the `EuroBioC2025_OSTA_Workshop` workflow by running

```{r}
# TODO
```

The relevant data are stored on Open Science Framework, and can be downloaded as delayed format cached on your laptop through `OSTA.data`:

```{r}
library(OSTA.data)
# TODO
```

### *R* / *Bioconductor* packages used

Bioconductor: Banksy, DropletUtils, ggspavis, OSTA.data, scuttle, spacexr, SpatialExperiment, SpotSweeper CRAN: dplyr, igraph, ggplot2, magick, patchwork, pheatmap, sf, tidyr, tidyverse

### Time outline

|            Activity          |   Time  |
|------------------------------|---------|
| Introduction                 |  5 mins |
| Initialization               |  3 mins |
| Data loading                 |  7 mins |
| Data processing              |  5 mins |
| Domain identification        | 10 mins |
| Deconvolution                | 15 mins |
| Conclusion                   |  5 mins |

## Workshop goals and objectives

The key steps to Visium HD data analysis are described in this workshop with diverse visualization of methods used. The goal is to allow beginner bioinformatician working on Visium HD data to become familiar with each of the steps involved, as well as completing a standard analysis pipeline from start to finish.

### Learning goals

-   learn how to analyse Visium HD data with cell segmentation output
-   understanding of key steps and methods for deriving biological insights 
-   aesthetic visualization of Visium HD data with existing R packages


### Learning objectives

-   read in cell segmentation `.geojson` data and format as a `sf` object
-   manipulate cell polygon coordinates with scaling factor and alignment with H&E image
-   read in count data and format as a `SpatialExperiment` object
-   perform spatially-aware quality control, clustering, marker gene identification
-   perform label transfering from single-cell reference data to spatial data via deconvolution
-   visualization of Visium HD data as polygons, of the cell centroids, and in reduced dimensions
-   checking concordance between cell typing results with heatmaps

## Introduction

As of June 2025, Visium HD enabled direct output of H&E segmented cell-level data from SpaceRanger v4. This automation skips the need of an additional segmentation tool, such as `bin2cell`. Since biologists often find having cellular units more intuitive when deriving insights, this output can attract more interests than bin-level data. However, out of cell boundary bins can still benefit many research topics in the tumor microenvironment. In this workflow, we will demonstrate some analysis steps at the cell-level using the same dataset, subsetted to the same region, as in [@sec-seq-workflow-visium-hd; @Oliveira2024].

## Dependencies {#sec-seq-workflow-visium-hd-cell-load-data}

```{r deps, message=FALSE, warning=FALSE}
library(Banksy)
library(DropletUtils)
library(ggplot2)
library(dplyr)
library(ggspavis)
library(igraph)
library(OSTA.data)
library(patchwork)
library(pheatmap)
library(scuttle)
library(spacexr)
library(SpatialExperiment)
library(SpotSweeper)
library(tidyr)

library(magick)
library(sf)
library(tidyverse)
```

## Load spatial data

Output files from SpaceRanger v4 are listed in `/segmented_outputs` folder, with the expected structure detailed in @sec-bkg-importing. We start with read in cell segmentation as a `sf` object, scaling factor `.json` file, and image file. We will use high resolution image, as it gives expected H&E details under segmented cell masks.

```{r load-other-vhdcell, message=FALSE, warning=FALSE}
# Read the GeoJSON file of cell segmentation
geo_data <- st_read("~/Desktop/cell_segmentations.geojson")
scalef <- rjson::fromJSON(file = "~/Desktop/scalefactors_json.json")
image <- image_read("~/Desktop/tissue_hires_image.png")
info <- image_info(image)
```

First, we need to scale the coordinates to match the high resolution image. In addition, the y-axis coordinates should be flipped, as R uses a different origin than the image coordinate system. While `ggspavis` manages this internally for Visium, we will proceed with a manual plotting approach for Visium HD polygons (rather than cell centroid x and y) to maintain clarity.

```{r scale-flip-polygons, message=FALSE, warning=FALSE}
# Multiply all coordinates by the scaling factor of high resolution image
st_geometry(geo_data) <- st_geometry(geo_data) * scalef$tissue_hires_scalef

# Flip axis by flip y coordinates with image height  - y axis
st_geometry(geo_data) <- st_sfc( 
    lapply(st_geometry(geo_data), function(geom) {
      coords <- st_coordinates(geom)
      coords[, 2] <- info$height - coords[, 2]
      st_polygon(list(matrix(coords[, 1:2], ncol = 2)))
    }),
    crs = st_crs(geo_data)
)
```

To keep runtimes low, our downstream analysis will be performed on a subset area of 16 µm bins (1/64 of the data coverage area).

```{r subset-polygons, message=FALSE, warning=FALSE}
# Get data coverage and subsetted bounding box
.rng <- \(geo_data) {
    cov <- st_bbox(geo_data)
    xs <- cov[c(1, 3)]; ys <- cov[c(2, 4)]
    x1 <- xs[1] + 4*(xs[2]-xs[1])/8; x2 <- xs[2] - 3*(xs[2]-xs[1])/8
    y1 <- ys[1] + 4*(ys[2]-ys[1])/8; y2 <- ys[2] - 3*(ys[2]-ys[1])/8
    list(box=c(x1, x2, y1, y2), cov=c(xs, ys))
}
geo_data_rng <- .rng(geo_data)

# Subset the GeoJSON to geometries that intersect with the bounding box
bbox_polygon <- st_as_sfc(st_bbox(geo_data_rng$box, crs = st_crs(geo_data)))
subset_data <- geo_data[st_intersects(geo_data, bbox_polygon, sparse = FALSE), ]
subset_data$cell_id <- as.character(subset_data$cell_id)
head(subset_data)
```

Here, as we have restricted the analysis to only 1/64 of the entire tissue coverage, the number of cells in the subset `sf` object should also be reduced by roughly this fraction, depending on tissue density.

```{r dim-polygons, message=FALSE, warning=FALSE}
nrow(geo_data)/nrow(subset_data)
```

Show subset region used for analysis:

```{r plt-subset-cov-box, fig.width=6, fig.height=4, message=FALSE}
.box <- \(rng, roi, lty, lwd, col) {
    data <- data.frame(xmin = rng[[roi]]["xmin"], xmax = rng[[roi]]["xmax"],
                       ymin = rng[[roi]]["ymin"], ymax = rng[[roi]]["ymax"])
    geom_rect(data = data, 
              aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
              color = col, fill = NA, linetype = lty, linewidth = lwd)
}
cov <- .box(rng = geo_data_rng, roi = "cov", lty = 2, lwd = 1/2, col = "#3b3b3b")
box <- .box(rng = geo_data_rng, roi = "box", lty = 4, lwd = 2/3, col = "black")

ggplot() + 
    annotation_raster(xmin = 0, xmax = info$width, ymin = 0, ymax = info$height, 
                      image) + 
    coord_fixed(xlim = c(0, info$width), ylim = c(0, info$height), 
                expand = FALSE) + cov + box + 
    ggtitle("Grey: data coverage; Black: subset region") + 
    theme_void() + theme(plot.title = element_text(hjust = 0.5))
```

Plot cell polygons overlay H&E image for this subset region (in black box). Here is an overview of the subset of tissue section that we will work on:

```{r plt-subset-he-polygon, fig.width=12, fig.height=4, message=FALSE}
#| code-fold: true
# Plot raster
p0 <- ggplot() +
    annotation_raster(image, 
                      xmin = 0, xmax = info$width, ymin = 0, ymax = info$height) +
    coord_fixed(expand = FALSE,
                xlim = geo_data_rng$box[1:2], ylim = geo_data_rng$box[3:4]) 

# Add sf layer
p <- p0 +
    geom_sf(data = subset_data, fill = NA, color = "white", size = 0.3) + 
    theme(axis.text = element_blank(),
          axis.ticks = element_blank())

# Just polygons
p1 <- ggplot(subset_data) +
    geom_sf(fill = "lightblue", color = "white") +
    theme_void()

# Show plots
(p0 + ggtitle("H&E region") |
    p + ggtitle("Overlayed with cell boundaries") |
    p1 + ggtitle("Cells")) +
    plot_layout(nrow=1) & theme(plot.title = element_text(hjust = 0.5))
```

## Load count data

Load the segmented cell-level count matrix from Visium HD data.

```{r load-counts-vhdcell, message=FALSE, warning=FALSE}
## VHD cell
countmat_file <- "~/Desktop/filtered_feature_cell_matrix.h5"
vhdcellsce <- DropletUtils::read10xCounts(countmat_file, col.names = TRUE)

# Fix colnames or cell_id in colData
vhdcellsce$cell_id <- colnames(vhdcellsce) <- 
    gsub("^cellid_0*([1-9][0-9]*|0)-1$", "\\1", colnames(vhdcellsce))

# Subset to region in sf
vhdcellsce <- vhdcellsce[, colnames(vhdcellsce) %in% subset_data$cell_id]

# Update gene and cell names
rownames(vhdcellsce) <- rowData(vhdcellsce)$Symbol
rownames(vhdcellsce) <- make.unique(rownames(vhdcellsce))
```

Obtain cell centroids from the subset `sf` object, and use them as spatial coordinates in the `SpatialExperiment` object.

```{r vhd-spacoords, message=FALSE, warning=FALSE}
centroids <- st_centroid(subset_data)
coords <- st_coordinates(centroids)

# sanity check on ordering of cell ids
all(centroids$cell_id == colnames(vhdcellsce))

head(coords)
```

Now, we construct a `SpatialExperiment` object from the cell-level count matrix.

```{r make-vhd, message=FALSE, warning=FALSE}
# construct a SPE, subsetted to region of interest
vhd <- SpatialExperiment(
    assays = list(counts = as(counts(vhdcellsce), "dgCMatrix")),
    rowData = rowData(vhdcellsce),
    colData = cbind(colData(vhdcellsce), coords),
    spatialCoordsNames = c("X", "Y")
)
vhd
```

## Quality control

Get cell-level QC metrics for visualization.

```{r qc-metrics-vhd, message=FALSE, warning=FALSE}
# Visualize library size to the polygons
sub <- list(mt = grep("^MT-", rownames(vhd)))
vhd <- addPerCellQCMetrics(vhd, subsets = sub)
vhd$log_sum <- log1p(vhd$sum)

CD <- as.data.frame(colData(vhd))
head(CD %>% select(cell_id, sum, log_sum, subsets_mt_percent))
```

Merge cell-level metadata into `sf` object with cell segmentation coordinates in the subset region.

```{r sf-join-qc-metric, message=FALSE, warning=FALSE}
subset_data <- subset_data %>% 
    left_join(CD %>% select(cell_id, log_sum, subsets_mt_percent))
head(subset_data)
```

We can visualize the metrics could be used for quality control.

```{r plt-qc-metrics, fig.width=8, fig.height=3.5, message=FALSE}
#| code-fold: true
# helper
plotPolygons <- function(objsf, annotate = "Cluster"){
    ggplot(objsf) +
      geom_sf(aes(fill = get(annotate)), color = "grey", linewidth = 0.1) +
      labs(fill = annotate) + 
      theme_void() + 
      theme(axis.text = element_blank(),
            axis.ticks = element_blank())
}

plotPolygons(subset_data, annotate = "log_sum") + 
    ggtitle("log library size") + 
    plotPolygons(subset_data, annotate = "subsets_mt_percent") + 
    ggtitle("% mitochondrial") + 
    plot_layout(nrow = 1) & 
    theme(plot.title = element_text(hjust = 0.5),
          legend.title=element_blank(),
          legend.key.width = unit(0.5, "lines"), 
          legend.key.height = unit(1, "lines")) & 
    scale_fill_gradientn(colors = pals::jet())
```

### SpotSweeper

SpotSweeper determine outliers based on low log-library size, few uniquely detected features, or high mitochondrial count fraction:

```{r qc-vhd-local}
vhd <- localOutliers(vhd, metric="sum", direction="lower", log=TRUE)
vhd <- localOutliers(vhd, metric="detected", direction="lower", log=TRUE)
vhd <- localOutliers(vhd, metric="subsets_mt_percent", direction="higher", log=TRUE)
vhd$discard <- 
    vhd$sum_outliers | 
    vhd$detected_outliers | 
    vhd$subsets_mt_percent_outliers
# tabulate number of cells retained vs. removed by any criterion 
table(vhd$discard)
```

We can visualize the local outliers identified by `SpotSweeper` in space. It is difficult to see as we only have few small cells identified as outliers per flag.

```{r qc-vhd-plt, fig.width=7.5, fig.height=2.5, message=FALSE}
#| code-fold: true
CD <- as.data.frame(colData(vhd)) 
subset_data <- subset_data %>%
    left_join(CD %>% select(cell_id, sum_outliers, detected_outliers, discard))

plotPolygons(subset_data, annotate="sum_outliers") + 
    ggtitle("low_lib_size") +
    plotPolygons(subset_data, annotate="detected_outliers") + 
    ggtitle("low_n_features") +
    plotPolygons(subset_data, annotate="discard") + 
    ggtitle("discard") +
    plot_layout(nrow=1, guides="collect") & 
    theme(plot.title=element_text(hjust=0.5),
          legend.key.size=unit(0, "lines")) &
    guides(col=guide_legend(override.aes=list(size=3))) & 
    scale_fill_manual("discard", values=c("white", "red"))
```

Filter out low quality cells from the `sf` and `SpatialExperiment` object.

```{r qc-filtering, message=FALSE, warning=FALSE}
subset_data_test <- subset_data[!vhd$discard, ]
vhd <- vhd[, !vhd$discard]

ncol(vhd); nrow(subset_data_test)
```

## Clustering

First, we identify highly variable genes (HVGs) in the target area: [Note that spatially variable genes (SVGs) could be used instead]{.aside}

```{r vhd-hvg}
vhd <- logNormCounts(vhd)
dec <- modelGeneVar(vhd)
hvg <- getTopHVGs(dec, n=3e3)
```

### Banksy

`r BiocStyle::Biocpkg("Banksy")` utilizes a pair of spatial kernels to capture gene expression variation, followed by dimension reduction and graph-based clustering to identify spatial domains.

```{r Banksy, message=FALSE}
# set seed for random number generation
# in order to make results reproducible
set.seed(112358)
# 'Banksy' parameter settings
k <- 8   # consider first order neighbors
l <- 0.2 # use little spatial information
a <- "logcounts"
xy <- c("array_row", "array_col")
# restrict to selected features
tmp <- vhd[hvg, ]
# compute spatially aware 'Banksy' PCs
tmp <- computeBanksy(tmp, assay_name=a, coord_names=xy, k_geom=k)
tmp <- runBanksyPCA(tmp, lambda=l, npcs=20)
reducedDim(vhd, "PCA") <- reducedDim(tmp)
## run UMAP (for visualization purposes only)
set.seed(112358)
vhd <- runUMAP(vhd, dimred="PCA")
# build cellular shared nearest-neighbor (SNN) graph
g <- buildSNNGraph(vhd, use.dimred="PCA", type="jaccard", k=20)
# cluster using Leiden community detection algorithm
k <- cluster_leiden(g, objective_function="modularity", resolution=1.2)
table(vhd$Banksy <- factor(k$membership))
```

Next, we can perform differential gene expresison (DGE) analysis to identify markers for each cluster; c.f. @sec-crs-signatures. We also compute the cluster-wise mean expression of selected markers:

```{r mgs}
# differential gene expression analysis
mgs <- findMarkers(vhd, groups=vhd$Banksy, direction="up")
# select for a few markers per cluster
top <- lapply(mgs, \(df) rownames(df)[df$Top <= 2])
top <- unique(unlist(top))
# average expression by clusters
pbs <- aggregateAcrossCells(vhd, 
    ids=vhd$Banksy, subset.row=top, 
    use.assay.type="logcounts", statistics="mean")
```

The marker genes in each cluster can be visualized with a heatmap:

```{r plt-mgs-hm, message=FALSE, fig.width=8.5, fig.height=3.4}
#| code-fold: true
# visualize averages z-scaled across clusters
pheatmap(
    mat=t(assay(pbs)), scale="column", breaks=seq(-2, 2, length=101), 
    cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```

Note that the visualization tools `library(ggspavis)` we previously developed for spot and cellular-level spatial transcriptomics technologies, e.g. Visium and Xenium, can also be used here, because we have calculated the cell centroid from the cell segmentation polygons. We can visualize the cell-level expression of selected markers in space:

```{r plt-mgs-xy, fig.width=10, fig.height=3, message=FALSE}
#| code-fold: true
gs <- c("MMP2", "PIGR", "IGHG1")
ps <- lapply(gs, \(.) 
             plotCoords(vhd, annotate=., point_size = 0.6, in_tissue = NULL,
                        assay_name="logcounts", y_reverse = FALSE))
wrap_plots(ps, nrow=1) & theme(
    legend.key.width=unit(0.5, "lines"),
    legend.key.height=unit(1, "lines")) &
    scale_color_gradientn(colors=rev(hcl.colors(9, "Rocket")))
```

```{r clus-merge-sf-umap-merge-clus, warning=FALSE, message=FALSE}
CD <- colData(vhd) %>% as.data.frame()
subset_data <- subset_data %>%
    left_join(CD %>% select(cell_id, Banksy))

umap <- reducedDim(vhd, "UMAP") %>% as.data.frame() %>%
    rownames_to_column("cell_id") %>%
    left_join(CD %>% select(cell_id, Banksy))
```

Now we can visualize clustering result spatially and on UMAP.

```{r clus-spa-umap, fig.width=7.3, fig.height=3.5, warning=FALSE, message=FALSE}
#| code-fold: true
# helper
plotCoordUmap <- function(subset_data, umap, annotate = "Cluster", 
                          legend.col = 2, pal = unname(pals::kelly())){
    pp <- plotPolygons(subset_data, annotate = annotate) + 
      scale_fill_manual(values = pal) + 
      theme(legend.position = "none",
            plot.title = element_text(hjust = 0.5)) + 
      ggtitle(annotate)
    
    pu <- ggplot(umap, aes(x = UMAP1, y = UMAP2)) +
      geom_point(aes(color = get(annotate)), size = 0.5) +
      scale_color_manual(values = pal) +
      labs(color = annotate) + 
      guides(col=guide_legend(override.aes=list(size=3), ncol = legend.col)) +
      theme_classic()
      
    return(pp | pu)
}

plotCoordUmap(subset_data, umap, annotate = "Banksy", legend.col = 2,
              pal = unname(pals::kelly()))
```

## Deconvolution

### Reference

As previously shown, we can label transfer from single-cell reference to spatial data using cell type deconvolution, assuming each data point contains around one cell type. First, we obtain the single-cell data as reference for deconvolution:

```{r load-sce, message=FALSE, warning=FALSE}
## SCE ref
id <- "Chromium_HumanColon_Oliveira"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir=td)

# read into 'SingleCellExperiment'
fs <- list.files(td, full.names=TRUE)
h5 <- grep("h5$", fs, value=TRUE)
sce <- read10xCounts(h5, col.names=TRUE)
# add cell metadata
csv <- grep("csv$", fs, value=TRUE)
cd <- read.csv(csv, row.names=1)
colData(sce)[names(cd)] <- cd[colnames(sce), ]
# use gene symbols as feature names
rownames(sce) <- make.unique(rowData(sce)$Symbol)
# exclude cells deemed to be of low-quality
sce <- sce[, sce$QCFilter == "Keep"]
# tabulate subpopulations
table(sce$Level1)
```

Down-sample to at most 4,000 cells per cluster for the single-cell reference.

```{r subset-sce, message=FALSE, warning=FALSE}
# (this is done only to keep runtime/memory low)
cs <- split(seq_len(ncol(sce)), sce$Level1)
cs <- lapply(cs, \(.) sample(., min(length(.), 4e3)))
ncol(.sce <- sce[, unlist(cs)])
```

### RCTD

We perform deconvolution with `doublet` mode for cell-level data, assuming at most 2 cell types per data point.

```{r rctd-run, warning=FALSE, message=FALSE}
rctd_data <- createRctd(vhd, .sce, cell_type_col="Level1")
(res <- runRctd(rctd_data, max_cores=4, rctd_mode="doublet"))
```

Check deconvolution class of cells. Note that some cells did not pass RCTD internal QC, so they have been filtered out prior to running deconvolution. You can check the number of genes and cells used to deconvolution with `dim(rctd_data$spatial_experiment)`.

```{r rctd-class, warning=FALSE, message=FALSE}
table(res$spot_class)
```

Majority of the cells are `singlet`, which confirms the build-in H&E guided segmentation in SpaceRanger. Then, we prepare deconvolution results for visualization.

```{r rctd-decon-result, warning=FALSE, message=FALSE}
ws <- assay(res, "weights")
ws <- data.frame(t(as.matrix(ws)))
res$cell_id <- colnames(res)
rctddf <- cbind(ws, colData(res) |> 
              as.data.frame() |> 
              select(cell_id, spot_class, first_type))
```

Merge cell-level deconvolution result into `sf` object that contains cell segmentation coordinates within the subset region.

```{r rctd-merge-sf, warning=FALSE, message=FALSE}
subset_data_plt <- subset_data %>% 
    right_join(rctddf)
colnames(subset_data_plt)
dim(subset_data_plt)
```

Deconvolution result per cell type.

```{r rctd-plt-decon-ct, fig.width=9.5, fig.height=8, warning=FALSE, message=FALSE}
#| code-fold: true
lapply(names(ws), \(.) 
       plotPolygons(subset_data_plt, annotate=.) + ggtitle(.)) |>
    wrap_plots(nrow=3, guides="collect") & theme(
      plot.title = element_text(hjust = 0.5),
      legend.title=element_blank(),
      legend.key.width=unit(0.5, "lines"),
      legend.key.height=unit(1, "lines")) &
    scale_fill_gradientn(colors=rev(hcl.colors(9, "Rocket")))
```

Next, we merge in deconvolution labels to the reduced dimension data frame. Only cells that passed RCTD quality control are kept.

```{r umap-merge-decon, warning=FALSE, message=FALSE}
umap <- umap %>% right_join(rctddf)
```

Visualize spatially and in reduced dimension by cell class and first cell type. For demonstration purpose only, we annotate each cell by their `first_type` from the deconvolution, the most likely cell type, regardless of their `spot_class` being `singlet`, `doublet_certain`, or `doublet_uncertain`.

```{r plt-coord-umap-decon, fig.width=8, fig.height=7, warning=FALSE, message=FALSE}
#| code-fold: true
plotCoordUmap(subset_data_plt, umap, annotate = "spot_class", legend.col = 1,
              pal = unname(pals::okabe())) / 
    plotCoordUmap(subset_data_plt, umap, annotate = "first_type", legend.col = 1,
                  pal = unname(pals::trubetskoy()))
```

## Downstream

### Concordance of clustering & deconvolution

Map over deconvolution result.

```{r rctd-map-vhd}
vhd$first_type <- rctddf$first_type[match(colnames(vhd), rctddf$cell_id)]
```

We can visualize the results spatially:

```{r plt-dec-clu-xy, fig.width=10, fig.height=4}
#| code-fold: true
plotCoords(vhd, annotate="Banksy", point_size=1.6, y_reverse = FALSE,
           in_tissue = NULL, pal=unname(pals::kelly())) + plot_spacer() +
plotCoords(vhd, annotate="first_type", point_size=1.6, y_reverse = FALSE,
           in_tissue = NULL, pal=unname(pals::trubetskoy())) + 
    plot_layout(nrow=1, widths=c(1, 0.05, 1)) & 
    theme(legend.key.size=unit(0, "lines"))
```

We observe that Banksy delineates tissue borders very well, while deconvolution provides direct biological insight into the underlying clusters. Their concordance can be visualized using a heatmap:

```{r plt-dec-clu-hm, fig.width=2.5, fig.height=3.5}
#| code-fold: true
fq <- prop.table(table(vhd$Banksy, vhd$first_type), 1)
pheatmap(fq, cellwidth=10, cellheight=10, treeheight_row=5, treeheight_col=5)
```
